Parameterization of the responses of subarctic European vegetation to key environmental variables for ozone risk assessment

### 2022-05-12 VegOzone.tex
Update bib and finish discussion and appendix

We found that 2018 was warmer, slightly drier, and brighter than usual, while 2019 was a rather average year at Svanhovd. According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record. Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the blocking high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. A total of 2079 forest fires were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden \citep{SOU2019}. Coincident peak \chem{\chi_{O_3}} is found in ozone monitoring data (Fig.~\ref{fig:data_svanvik_2018_2019}) in July and ozone was more enhanced in 2018 than 2019. It is very likely that elevated ozone, like during the 2003 drought period \citep{JGR:Solberg2018}, was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions. As such events may become more frequent in the future, conditions at Svanhovd for the growing seasons 2018 could serve as a reference for probable future conditions in northern Fennoscandia. Conditions in 2019 represent the present climate.

Taking a presumed adaption of vegetation to local climate into consideration, we derived localized temperature and light response functions for modeling $\mathrm{POD_1}$ and biomass reduction with the $\mathrm{DO_3SE}$ model. Our key findings are that vegetation that is more sensitive to warm temperatures would be less effected by coincident ozone and heat stress because stomata will close to prevent water loss. On the other hand, a warmer climate and hence prolonged GS would negatively affect vegetation adapted to the subarctic. It could be essential for ozone risk assessments in the subarctic to define plant physiology and traits properly. For comparison, $\mathrm{DO_3SE}$ model derived dry deposition velocities $V_\mathrm{d}$ for coniferous trees with any parameterization are comparable with observations by \citet{BER:Keronen2003} (refer to supplementary material) but might show slightly too high $V_\mathrm{d}$ in spring. As perennial grassland and deciduous trees are more strongly affected by our localization (increase in frequency of maximum $V_\mathrm{d}$ throughout the GS), flux measurements over these vegetation types in a subarctic climate would be needed to evaluate our proposed method of localization.

We found that the difference in biomass reduction compared to Boreal parmeterizations \citep{ICP:MappingManual2017} could amount to up to $55\,\unit{\%}$ (deciduous trees). Considering perennial grassland with predicted above ground biomass reduction due to ozone of $12-14\,\unit{\%}$ in 2018/2019 ($29-43\,\unit{\%}$ difference). From an agro-economical point of view in our focus area, where the grass is cut only once a year, this predicted loss of biomass is large and could be more severe than in more productive areas in other parts of Europe. Therefore, the CL for protection of vegetation of $10\,\unit{\%}$ may need adaption.

Along the conventional ozone monitoring at Svanhovd, we had installed a bio-monitoring with ozone sensitive species. Qualitatively, this showed visible ozone damage on clover and tobacco in 2018. In contrast, no such damage was found on the clovers in 2019. This might indicate a higher ozone damage risk under warmer climate conditions for these specific plant species in a subarctic climate. However, visible damage is not always correlated with a reduction in biomass \citep[and citations therein]{CRG:Felzer2007}
and our $\mathrm{DO_3SE}$ modeling results are in contrast to these observations. Visible damage caused by peak ozone uptake and specific dependency of ozone sensibility on phenology stage are not accounted for in the $\mathrm{DO_3SE}$ model. In this regard, open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations attributed a reduction in biomass and reproduction with visible damage explicitly to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. In particular, forbs and perennial grassland are more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. 
That might explain why we observed damage in the ozone garden to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.

Based on a MODIS GPP product \citep{MODIS_PSN}, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The determined start of growing season in 2018 ($\unit{doy}~122$) and 2019 ($\unit{doy}~106$) lies within the range given by observations in Fennoscandia \citep{TB:Kolari2007,IVL:Karlsson2018}. We determined the start of the growing season for deciduous trees based on gridded observational temperature and for perennial grassland on snow depth data. For perennial grassland, we assumed a one-month latency period after snowmelt that is supported by observational data from Rovaniemi (Finland) \citep{FCR:Korhonen2018}.

TODO:
We found that soil water potential under 2018/19 meteorological conditions was negligible when considering the bespoke growing season. We have not assessed the effect of the plants' response to VPD in detail, but our climatological data indicated that conditions at Svanhovd are shifting towards a more VPD limited regime.

In this regard, vegetation might become more exposed to higher ozone concentrations in early phenological stages due to an overlap with ozone spring peak \citep{EP:Karlsson2007}. However, the decline of this ozone spring peak is partly caused by the uptake of vegetation. It, therefore, remains unclear whether an earlier start of the growing season will increase the exposure of vegetation to ozone or lead to compensation due to an earlier decline of the peak ozone in the future. \citet{ESPR:Hayes2021} pointed out that the highest susceptibility of vegetation to future ozone variability is likely to occur in summer when vegetation is most productive. The actual resilience of northern Fennoscandia plant communities to climate change strongly depends on both the range of existing acclimations as well as the actual type of acclimation. This means vegetation that is acclimated to cold temperatures and little to heat- and drought stress might suffer more strongly from a more frequent occurence of heatwaves than plants that have a higher temperature tolerance.

Following the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and found substantial reductions for deciduous trees ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) in both years. These reductions are higher than for the default parameterizations, $11.2\,\unit{\%}$ (deciduous) and $9.6\,\unit{\%}$ (perennial grassland). For coniferous trees, the estimated reduction in biomass differs insignificantly between our bespoke parameterizations ($2.5-3\,\unit{\%}$) and the default ($2.31\,\unit{\%}$). This indicates a higher modeled robustness to ozone damage in coniferous trees. All these results have to be treated with care because the biomass reduction functions were established based on studies from less extreme climates. In particular, CLs for deciduous and coniferous trees were breached in all years and for all parameterizations. For perennial grassland, the CL was not breached if a bespoke GS was assumed, but the mapping manual defined CL of $10\,\unit{\%}$ loss in biomass might not be transferable to subarctic regions with considerable constraints on productivity leading to more severe economical consequences.

Beyond the risk assessment of ozone-induced damage, our bespoke temperature response functions, once verified by observations, can have important implications on land-surface modeling in global models, where problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. We pointed out, concerning default perennial grassland parameterization, that these perennial grasslands did not show any substantial stomatal conductance in 2019 characterized as a normal year. In terms of GPP this would mean unrealistically low carbon uptake. Automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.

@@

\begin{figure*}[th]
  \includegraphics[width=12cm]{greening_season_change_Svanvik}
  \caption{Estimated shift and prolongation of growing season at Svanhovd over the past 6 decades based on data from \citet{SeNorge}.}
  \label{fig:greening_season_change_Svanvik}
\end{figure*}

@@

\subsection{Initial $\mathrm{DO_3SE}$ modeling results with mapping manual parameters}
To demonstrate the necessary acclimation of the mapping manual parameters, we show the initial results of stomatal conductance at leaf-level ($G_\mathrm{sto}^\mathrm{leaf}$) and $\mathrm{POD_y}$ over \unit{doy} for both years and all PFTs (Fig.~\ref{fig:pody_mm_composit}). Observed \chem{\chi_{O_3}} is also indicated. From Fig.~\ref{fig:pody_mm_composit}f) it is apparent that the mapping manual parameterized grassland would not have been able to grow in 2019.

@@

\subsection{Bespoke parameterizations}

To assess the $G_\mathrm{start}$  and $G_\mathrm{end}$ of the growing season for coniferous trees at Svanhovd in 2018/19, we used the net photosynthesis product ($A_\mathrm{net}$) of MODIS AQUA/TERRA over a $1\times 1\,\unit{km}$ area centered at Svanhovd. MODIS data indicate a higher photosynthetic activity in 2018 than in 2019. As shown in Fig.~\ref{fig:modis_gpp} we fitted a second order polynomial function of the general form
%
\begin{equation}
A_\mathrm{net}(t) =  -m_0\cdot(t-m_1)^2+m_2,
\end{equation}
%
with the form parameters $m_i$ through the data. Numerically, we retrieved the roots as $G_\mathrm{start}$ 122/106 and $G_\mathrm{end}$ 261/274 for 2018 and 2019, respectively.

\begin{figure}[th]
  %\includegraphics[width=8.3cm]{figB2}
  \includegraphics[width=8.3cm]{modis_gpp}
  \caption{Estimated $G_\mathrm{start}$ and $G_\mathrm{end}$ of growing season for coniferous trees from MODIS Aqua/Terra gross primary productivity (GPP) product. A $1\times 1\,\unit{km}$ area around Svanhovd was selected. Daily averaged data for both 2018 and 2019 has been fitted with a quadratic polynomial function. The numerically computed root yields: BGS \unit{doy}~122/106 and EGS \unit{doy}~261/274 for 2018/19, respectively.}
  \label{fig:modis_gpp}
\end{figure}

As described in Section~\ref{subsec:do3se_parameters}, we developed bespoke parameterizations for natural and semi-natural vegetation at Svanhovd. Here, we show the temperature response and light response functions for coniferous and deciduous trees (Figs.~\ref{fig:f_temp_spruce}-\ref{fig:f_temp_birch}).

\begin{figure}[t]
  \centering
  (a)\\
  \includegraphics[width=8.3cm]{javis_funcs_temp_hist_coniferous}\\
  (b)\\
  \includegraphics[width=8.3cm]{javis_funcs_light_hist_coniferous.pdf}
  %\includegraphics[width=8.3cm]{figB3}
\caption{Construction of bespoke response functions for coniferous trees. (a) $f_\mathrm{T}$ and (b) $f_\mathrm{flight}$ are shown together with underlying $T_\mathrm{air}$ and $Q_0$ climatologies (probability density function - PDF), respectively. Original mapping manual parameterization is shown in comparison as solid line. Note that $Q_0$ has been truncated to $0.006$. PPFD0.8 and PPFD1.2 refer to $\alpha$ values increasing/decreasing PPFD at $f_\mathrm{light}=0.5$ by $\pm 20\,\%$, respectively.}
\label{fig:f_temp_spruce}
\end{figure}

>>>>

We found that 2018 was warmer, slightly drier, and brighter than usual, while 2019 was a rather average year at Svanhovd. According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record. Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the blocking high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe. A total of 2079 forest fires were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep[last accessed April 2020]{DSB2019}. In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden \citep{SOU2019}. We found coincident peak \chem{\chi_{O_3}} in ozone monitoring data at Svanhovd in July. Overall, ozone was more enhanced in 2018 than 2019. It is very likely that elevated ozone, like during the 2003 drought period \citep{JGR:Solberg2018}, was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions. As such events may become more frequent in the future, conditions at Svanhovd for the growing seasons 2018 could serve as a reference for probable future conditions in northern Fennoscandia and arising ozone-induced damage risk. Conditions in 2019 represent the present climate.

Taking a presumed adaption of vegetation to local climate into consideration, we derived localized Jarvis-type temperature and light response functions for modeling $\mathrm{POD_1}$ and biomass reduction with the $\mathrm{DO_3SE}$ model. Our key findings are that vegetation that is more sensitive to warm temperatures would be less effected by coincident ozone and heat stress because stomata will close to prevent water loss whereas vegetation adapted to warmer climate zones could be under higher ozone damage risk under these conditions. On the other hand, a warmer climate and hence prolonged GS would negatively affect any vegetation using the $\mathrm{POD_y}$ metric. It is essential for ozone risk assessments in the subarctic to define plant physiology and traits such as phenological states properly. As no independent ozone flux measurements are available at Svanhovd, we could not fully validate our methodology at this point. 
A comparison of $\mathrm{DO_3SE}$ model derived dry deposition velocities $V_\mathrm{d}$ for coniferous trees with comprehensive records over Pine forests from Hyytilälä, Finland \citep{BER:Keronen2003} (refer to supplementary material) indicate a slightly too high $V_\mathrm{d}$ in spring for our most extreme localization (subarctic). No comparable data is available for northern Fennoscandia or deciduous tress and perennial grassland. Perennial grassland and deciduous trees displayed strong susceptibility to our localization methodology (increase in frequency of maximum $V_\mathrm{d}$ throughout the GS, see supplementary material) but are within the range given in the literature \citep[e.g.][]{JGR:Jacob1992,AE:Padro1996,RG:Clifton2020}. Independent flux measurements over these vegetation types in a subarctic climate are thus crucial for evaluation and validation.

Along the conventional ozone monitoring at Svanhovd, we had installed a bio-monitoring with ozone sensitive species. Qualitatively, this showed visible ozone damage on clover and tobacco in 2018. In contrast, no such damage was found on the clovers in 2019. This might indicate a higher ozone damage risk under warmer climate conditions for these specific plant species in a subarctic climate. However, visible damage is not always correlated with a reduction in biomass \citep[and citations therein]{CRG:Felzer2007}. Also, our $\mathrm{DO_3SE}$ modeling results are in contrast to these observations at Svanhovd. Any visible damage caused by peak ozone flux through stomata or a specific dependency of plant sensibility to ozone on phenology stage are not accounted for in the $\mathrm{DO_3SE}$ model. In this regard, open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations attributed a reduction in biomass and reproduction with visible damage explicitly to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. Forbs and perennial grassland are more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. 
That might explain why we observed damage in the ozone garden to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.

Based on the MODIS GPP product \citep{MODIS_PSN}, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The determined start of growing season in 2018 ($\unit{doy}~122$) and 2019 ($\unit{doy}~106$) lies within the range given by observations in Fennoscandia \citep{TB:Kolari2007,IVL:Karlsson2018}. We determined the start of the growing season for deciduous trees using growing-degree-days above $5\,\unit{^\circ C}$ on gridded observational temperature. The results coincide with the first snow free days in our target region.
For the period $1994-2003$ \citet{IJB:Shutova2006} reported an observed average start of the GS for mountain birch at Svanhovd at $\unit{doy}~154$ and end of GS at $\unit{doy}~ 229$. These dates are well in line with our estimated average start and end of GS for the period $1960-2020$ (Fig.~\ref{fig:greening_season_change_Svanvik}).
Our estimates for perennial grassland, are supported by observational data from Rovaniemi (Finland) \citep{FCR:Korhonen2018}. The end of GS inferred from the zero-crossing of MODIS observed GPP is in line with independent estimates from the gridded observational temperature. Our method differs from NDVI based determination of GS \citep{IGARSS:Hogda2001,IJB:Karlsen2007, RS:Hogda2013}.
The \citet{IJB:Shutova2006} data indicate an earlier end of GS for birch compared to 2018/19. Therefore, $\mathrm{POD_1}$ for deciduous trees could be underestimated by $1.8\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $3.0\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019, respectively if a climatological GS is used. For perennial grassland the resulting underestimation would amount to $2.0\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $2.3\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. This demonstrates the importance of an accurate determination of GS for $\mathrm{POD_1}$ based risk assessments.

As shown by \citet{ACP:Bueker2012}, soil water potential is a very important factor for modeling stomatal conductance in the $\mathrm{DO_3SE}$ model. Using our best estimate of the soil type at Svanhovd, we found that SWP under 2018/19 meteorological conditions was mainly negligible when considering the localized growing season. We have not assessed the effect of the plants' response to VPD in detail, but the climatological data indicated that conditions at Svanhovd are shifting towards a more VPD limited regime that could limit $\mathrm{POD_y}$ under drought stress.

Due to a climate change-induced convergence of start of GS with the ozone spring peak \citep{EP:Karlsson2007}, vegetation might become more exposed to higher ozone concentrations in early phenological stages. However, the decline of this ozone spring peak is partly caused by dry deposition of ozone to vegetation itself. It, therefore, remains unclear whether an earlier start of the growing season will increase the exposure of vegetation to ozone or lead to compensation due to an earlier decline of the peak ozone in the future. \citet{ESPR:Hayes2021} pointed out that the highest susceptibility of vegetation to future ozone variability is likely to occur in summer when vegetation is most productive. The actual resilience of northern Fennoscandia plant communities to climate change strongly depends on both the range of acclimation and the type of acclimation. This means, vegetation that is acclimated to cold temperatures and less to heat- and drought stress might suffer more strongly from a more frequent occurrence of heatwaves than plants that have a higher temperature tolerance, but our results point to the opposite concerning the ozone stress.

We found that the difference in biomass reduction compared to Boreal parmeterizations \citep{ICP:MappingManual2017} could amount to $55\,\unit{\%}$ in deciduous trees and $29-43\,\unit{\%}$ in perennial grassland. But the biomass reduction functions were established based on studies from less extreme climates, hence, these results ought to be treated with care. In particular, CLs for deciduous and coniferous trees were breached in both years and for all parameterizations. For perennial grassland, the CL was not breached if a localized GS was assumed, but the defined CL of $10\,\unit{\%}$ loss in biomass might not be transferable to subarctic regions with considerable constraints on productivity leading to more severe economic consequences.

Beyond the risk assessment of ozone-induced damage, our localized Jarvis-type temperature response functions, once verified by observations, can have important implications on land-surface modeling still using such parameterizations in global models. Problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. A verification of our proposed methodology by independent flux and dry deposition measurements over birch forest and perennial grassland would be necessary to establish a climate data based localization of stomatal conductance parameterizations.

@@

\begin{figure*}[th]
  \includegraphics[width=12cm]{greening_season_change_Svanvik}
  \caption{Estimated shift and prolongation of growing season at Svanhovd over the past 6 decades based on gridded obervational temperature data from \citet{SeNorge}.}
  \label{fig:greening_season_change_Svanvik}
\end{figure*}

@@

\subsection{Initial $\mathrm{DO_3SE}$ modeling results with mapping manual parameters}
To demonstrate the necessary localization of the mapping manual parameters, we show the initial results of stomatal conductance at leaf-level ($G_\mathrm{sto}^\mathrm{leaf}$) and $\mathrm{POD_y}$ over \unit{doy} for both years and all PFTs (Fig.~\ref{fig:pody_mm_composit}). Observed \chem{\chi_{O_3}} is also indicated.

@@

\subsection{Localized parameterizations}

To assess the $G_\mathrm{start}$  and $G_\mathrm{end}$ of the growing season for coniferous trees at Svanhovd in 2018/19, we used the gross photosynthesis product (GPP) of MODIS AQUA/TERRA over a $1\times 1\,\unit{km}$ area centered at Svanhovd. MODIS data indicate a higher photosynthetic activity in 2018 than in 2019. As shown in Fig.~\ref{fig:modis_gpp}, we fitted a second order polynomial function of the general form
%
\begin{equation}
A_\mathrm{net}(t) =  -m_0\cdot(t-m_1)^2+m_2,
\end{equation}
%
with the form parameters $m_i$ to the data. Numerically, we retrieved the roots as $G_\mathrm{start}$ 122/106 and $G_\mathrm{end}$ 261/274 for 2018 and 2019, respectively.

\begin{figure}[th]
  %\includegraphics[width=8.3cm]{figB2}
  \includegraphics[width=8.3cm]{modis_gpp}
  \caption{Estimated $G_\mathrm{start}$ and $G_\mathrm{end}$ of growing season for coniferous trees from MODIS Aqua/Terra gross primary productivity (GPP) product. A $1\times 1\,\unit{km}$ area around Svanhovd was selected. Daily averaged data for both 2018 and 2019 has been fitted with a quadratic polynomial function.}
  \label{fig:modis_gpp}
\end{figure}

Here, we show the temperature and light response functions for coniferous and deciduous trees (Figs.~\ref{fig:f_temp_spruce}-\ref{fig:f_temp_birch}).

\begin{figure}[t]
  \centering
  (a)\\
  \includegraphics[width=8.3cm]{javis_funcs_temp_hist_coniferous}\\
  (b)\\
  \includegraphics[width=8.3cm]{javis_funcs_light_hist_coniferous.pdf}
  %\includegraphics[width=8.3cm]{figB3}
\caption{Construction of localized response functions for coniferous trees. (a) $f_\mathrm{T}$ and (b) $f_\mathrm{flight}$ are shown together with underlying $T_\mathrm{air}$ and $Q_0$ climatologies (probability density function - PDF), respectively. Original mapping manual parameterization is shown in comparison as solid line. Note that the y-axis has been truncated to $0.006$. PPFD0.8 and PPFD1.2 refer to $\alpha$ values increasing/decreasing PPFD at $f_\mathrm{light}=0.5$ by $\pm 20\,\%$, respectively.}
\label{fig:f_temp_spruce}
\end{figure}
